---
title: "panel_fe_mem"
output: html_document
date: "2026-03-15"
---

## 0. Packages

```{r message=FALSE, warning=FALSE}
# install.packages(c(
#   "sf", "terra", "osmdata", "dplyr", "sfnetworks", "tidygraph",
#   "MASS", "ggplot2", "tidyterra", "lubridate", "tidyr", "fixest",
#   "h3jsr", "spdep", "spatialreg", "purrr"
# ))

library(modelsummary)
library(stringr)
library(purrr)
library(pscl)
library(sf)
library(terra)
library(osmdata)
library(dplyr)
library(sfnetworks)
library(tidygraph)
library(MASS)
library(ggplot2)
library(tidyterra)
library(lubridate)
library(tidyr)
library(fixest)
library(h3jsr)
library(spdep)
library(spatialreg)
library(purrr)
```


## 1. Files and input data

```{r}
GHSL_FILE  <- "/Users/jachym.bielesz/diplomka/ghsl/GHS_BUILT_S_E2018_GLOBE_R2023A_54009_10_V1_0_R4_C20.tif"
CRIME_FILE <- "/Users/jachym.bielesz/diplomka/kriminalita_geojson/crimes_prague.geojson"

r <- rast(GHSL_FILE)
names(r) <- "ghsl"

crimes <- st_read(CRIME_FILE, quiet = TRUE)

if (is.na(st_crs(crimes))) {
  st_crs(crimes) <- 4326
}
```

## 2. Crime time variables

```{r}
crimes <- crimes |>
  mutate(
    crime_time = ymd_hms(date),
    year       = year(crime_time),
    month      = month(crime_time),
    year_month = floor_date(crime_time, "month")
  ) |>
  filter(!is.na(year), !is.na(month))
```

## 3. Prague boundary

```{r}
prague_poly <- getbb("Praha, Czech Republic", format_out = "polygon")

prague <- st_sfc(
  lapply(prague_poly, function(feature) {
    coords <- as.matrix(feature[[1]])
    storage.mode(coords) <- "double"
    st_polygon(list(coords))
  }),
  crs = 4326
) |>
  st_as_sf() |>
  summarise() |>
  st_make_valid()

prague_wgs <- st_transform(prague, 4326)
```

```{r}
ggplot() +
  geom_sf(data = prague_wgs, fill = "lightblue", color = "black", alpha = 0.4) +
  theme_minimal() +
  labs(title = "Prague boundary")
```

## 4. Filter crimes to Prague

```{r}
crimes_clean <- crimes |>
  mutate(
    types_list = str_split(types, ";"),
  )

crimes_clean$types_list <- lapply(crimes_clean$types_list, as.numeric)
```


```{r}
crimes_prague <- st_intersection(crimes_clean, prague_wgs) |>
  filter(year > 2021) #|>
  #filter(map_lgl(types_list, ~ any(.x >= 300 & .x < 400, na.rm = TRUE))) # sexualni zlociny
```

```{r}
ggplot() +
  geom_sf(data = prague_wgs, fill = NA, color = "grey60", linewidth = 0.4) +
  geom_sf(
    data = crimes_prague,
    color = "red", size = 0.4, alpha = 0.4
  ) +
  theme_void() +
  labs(title = "Recorded crimes in Prague, 2020")
```

## 5. H3 grid over Prague

```{r}
res_h3 <- 8

h3_ids <- polygon_to_cells(prague_wgs, res = res_h3, simple = TRUE)
h3_ids <- h3_ids[!is.na(h3_ids)]

hex_sf <- cell_to_polygon(h3_ids, simple = FALSE) |>
  mutate(id = seq_len(n()))

hex_54009 <- st_transform(hex_sf, crs(r))
crimes_54009 <- st_transform(crimes_prague, crs(r))
```

```{r}
## Map of H3 grid over Prague

grid_plot <- st_transform(hex_sf, st_crs(prague_wgs))

ggplot() +
  geom_sf(data = grid_plot,
          fill = NA,
          color = "steelblue",
          linewidth = 0.2) +
  geom_sf(data = prague_wgs,
          fill = NA,
          color = "black",
          linewidth = 0.6) +
  coord_sf(expand = FALSE) +
  theme_void() +
  labs(title = paste("H3 grid over Prague (resolution =", res_h3, ")"))
```


## 6. Crime counts per cell and year

```{r}
join_crime <- st_join(crimes_54009, hex_54009, join = st_within)

crime_counts_month <- join_crime |>
  st_drop_geometry() |>
  group_by(id = id.y, year, month) |>
  summarise(crime_count = n(), .groups = "drop")

ids <- hex_54009$id

time_index <- crimes_54009 |>
  st_drop_geometry() |>
  distinct(year, month) |>
  arrange(year, month)

grid_time_panel <- expand_grid(
  id = ids,
  time_index
)

crime_counts_month <- grid_time_panel |>
  left_join(crime_counts_month, by = c("id", "year", "month")) |>
  mutate(crime_count = replace_na(crime_count, 0L))

grid_crime <- hex_54009 |>
  left_join(crime_counts_month, by = "id") |>
  mutate(crime_count = replace_na(crime_count, 0L))
```

## 7. GHSL built-up intensity and WORLDPOP population

```{r}
vals <- terra::extract(r, vect(grid_crime), fun = mean, na.rm = TRUE)
grid_crime$ghsl_mean <- vals$ghsl
```

```{r}
POP_FILE <- "/Users/jachym.bielesz/diplomka/worldpop/cze_pop_2015_CN_100m_R2025A_v1.tif"

pop_r <- rast(POP_FILE)
names(pop_r) <- "population"

hex_pop_crs <- st_transform(hex_54009, crs(pop_r))

pop_vals <- terra::extract(
  pop_r,
  vect(hex_pop_crs),
  fun = sum,
  na.rm = TRUE
)

hex_population <- hex_54009 |>
  mutate(
    population = pop_vals$population,
    population = replace_na(population, 0)
  )

population_vars <- hex_population |>
  st_drop_geometry() |>
  select(id, population)
```



## 8. Street network from OSM

```{r}
bb <- st_bbox(prague_wgs)

car_highways <- c(
  "motorway", "trunk", "primary", "secondary", "tertiary",
  "motorway_link", "trunk_link", "primary_link",
  "secondary_link", "tertiary_link", "residential"
)

q_streets <- opq(bbox = bb) |>
  add_osm_feature(key = "highway", value = car_highways)

osm <- osmdata_sf(q_streets)
streets_raw <- osm$osm_lines
rm(osm); gc()
```

```{r}
streets_prague <- streets_raw |>
  st_make_valid()
```

```{r}
streets_clean <- streets_prague |>
  filter(st_geometry_type(geometry) %in% c("LINESTRING", "MULTILINESTRING")) |>
  st_cast("LINESTRING") |>
  filter(!st_is_empty(geometry))
```

## 9. Street network centrality

```{r}
net_prague <- as_sfnetwork(
  streets_clean,
  directed = FALSE,
  edges_as_lines = TRUE
)

net_prague_m <- st_transform(net_prague, 3857)

net_prague_m <- net_prague_m |>
  activate(nodes) |>
  mutate(
    deg  = centrality_degree(mode = "all"),
    betw = centrality_betweenness(),
    clos = centrality_closeness(),
    eig  = centrality_eigen()
  )

nodes_sf <- st_as_sf(net_prague_m, "nodes")
```

## 10. Aggregate centralities to grid

```{r}
grid_3857 <- st_transform(grid_crime, 3857)

join_nodes <- st_join(grid_3857, nodes_sf, join = st_contains)

cent_grid_nodes <- join_nodes |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    deg_mean   = ifelse(all(is.na(deg)), 0, mean(deg, na.rm = TRUE)),
    betw_mean  = ifelse(all(is.na(betw)), 0, mean(betw, na.rm = TRUE)),
    clos_mean  = ifelse(all(is.na(clos)), 0, mean(clos, na.rm = TRUE)),
    eig_mean   = ifelse(all(is.na(eig)), 0, mean(eig, na.rm = TRUE)),
    .groups    = "drop"
  )

grid_full <- grid_crime |>
  left_join(cent_grid_nodes, by = "id")
```

## 11. Metro stations

```{r}
q_metro <- opq(bbox = bb) |>
  add_osm_feature(key = "railway", value = c("station", "halt", "subway_entrance"))

osm_metro <- osmdata_sf(q_metro)
metro_pts <- osm_metro$osm_points |>
  st_make_valid()

metro_54009 <- st_transform(metro_pts, crs(grid_crime))

join_metro <- st_join(
  grid_crime,
  metro_54009,
  join = st_is_within_distance,
  dist = 500
)

metro_grid <- join_metro |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    has_metro_500 = as.integer(any(!is.na(osm_id))),
    .groups = "drop"
  )
```

## 12. Pubs, bars, restaurants, cafés

```{r}
q_pubs <- opq(bbox = bb) |>
  add_osm_feature(
    key = "amenity",
    value = c("pub", "bar", "restaurant", "cafe")
  )

osm_pubs <- osmdata_sf(q_pubs)
pub_pts <- osm_pubs$osm_points |>
  st_make_valid()

pubs_54009 <- st_transform(pub_pts, crs(grid_crime))

join_pubs <- st_join(
  grid_crime,
  pubs_54009,
  join = st_is_within_distance,
  dist = 500
)

pubs_grid <- join_pubs |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    n_pubs_500 = sum(!is.na(osm_id)),
    .groups = "drop"
  )
```

## 13. Merge static regressors

```{r}
grid_full <- grid_full |>
  left_join(pubs_grid,  by = "id") |>
  left_join(metro_grid, by = "id")
 grid_full <- grid_full %>% mutate(
    has_metro_500 = ifelse(is.na(has_metro_500), 0L, has_metro_500),
    n_pubs_500    = ifelse(is.na(n_pubs_500), 0L, n_pubs_500)
  )

static_vars <- grid_full |>
  st_drop_geometry() |>
  select(
    id,
    ghsl_mean,
    betw_mean,
    clos_mean,
    n_pubs_500,
    has_metro_500,
    eig_mean,
    deg_mean
  ) |>
  left_join(population_vars, by = "id") |>
  mutate(
    population = replace_na(population, 0),
    pop_offset = log(population + 1)
  )
```

## 14. Prepare panel data

```{r}
ids <- sort(unique(static_vars$id))

time_index <- crime_counts_month |>
  distinct(year, month) |>
  arrange(year, month)

panel_base <- expand_grid(
  id = ids,
  time_index
)

static_vars <- distinct(static_vars)

panel_df <- panel_base |>
  left_join(crime_counts_month, by = c("id", "year", "month")) |>
  mutate(crime_count = replace_na(crime_count, 0L)) |>
  left_join(static_vars, by = "id")

panel_df_model <- panel_df
```

```{r}
panel_df_model <- panel_df #|>
#  filter(
#    !is.na(ghsl_mean),
#    !is.na(betw_mean),
#    !is.na(clos_mean),
#    !is.na(n_pubs_500),
#    !is.na(eig_mean),
#    !is.na(deg_mean),
#    !is.na(has_metro_500)
#  )
```

## 15. Basic diagnostics

```{r}
mean(panel_df_model$crime_count)
var(panel_df_model$crime_count)
mean(panel_df_model$crime_count == 0)
```

## 16. Spatial weights for MEM

```{r}
grid_mem <- grid_full |>
  select(id) |>
  distinct() |>
  arrange(id)

nb <- poly2nb(grid_mem, queen = TRUE)
table(card(nb))

lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
```


## 17. Cross-section for MEM selection

```{r}
mem_select_df <- panel_df_model |>
  group_by(id) |>
  summarise(
    crime_avg      = mean(crime_count, na.rm = TRUE),
    ghsl_mean      = first(ghsl_mean),
    betw_mean      = first(betw_mean),
    clos_mean      = first(clos_mean),
    n_pubs_500     = first(n_pubs_500),
    has_metro_500  = first(has_metro_500),
    eig_mean       = first(eig_mean),
    deg_mean       = first(deg_mean),
    population     = first(population),
    pop_offset     = first(pop_offset),
    .groups        = "drop"
  ) |>
  arrange(id)

stopifnot(all(mem_select_df$id == grid_mem$id))
```


## 18. Select MEM eigenvectors

```{r echo=TRUE, message=FALSE, warning=FALSE}
# run MEM selection
#me_obj <- ME(
#  crime_avg ~ 1,
#  data = mem_select_df,
#  listw = lw,
#  family = poisson(link = "log"),
#  alpha = 0.05
#)

```

## 19. Add MEM variables to panel



```{r}
#if (is.null(me_obj$vectors) || ncol(as.data.frame(me_obj$vectors)) == 0) {
#  mem_vars_df <- mem_select_df |>
#    select(id)
#} else {
#  mem_mat <- as.data.frame(me_obj$vectors)
#  names(mem_mat) <- paste0("mem_", seq_len(ncol(mem_mat)))
#
#  mem_vars_df <- bind_cols(
#    mem_select_df |> select(id),
#    mem_mat
#  )
#}
#
#panel_df_mem <- panel_df_model |>
#  left_join(mem_vars_df, by = "id")

#mem_names <- names(panel_df_mem)[grepl("^mem_", names(panel_df_mem))]
#mem_names
```

```{r}
## 1️⃣ Agregace (SUM místo mean)
mem_select_df <- panel_df_model |>
  group_by(id) |>
  summarise(
    crime_sum = sum(crime_count, na.rm = TRUE),
    ghsl_mean = first(ghsl_mean),
    betw_mean = first(betw_mean),
    clos_mean = first(clos_mean),
    n_pubs_500 = first(n_pubs_500),
    has_metro_500 = first(has_metro_500),
    eig_mean = first(eig_mean),
    deg_mean = first(deg_mean),
    .groups = "drop"
  )

## 2️⃣ Odstranit izolované buňky
keep <- card(nb) > 0

mem_select_df_sub <- mem_select_df[keep, ]
nb_sub <- subset(nb, subset = keep)
lw_sub <- nb2listw(nb_sub, style = "W", zero.policy = TRUE)

## 3️⃣ MEM selection
me_obj <- ME(
  crime_sum ~ 1,
  data = mem_select_df_sub,
  listw = lw_sub,
  family = poisson(link = "log"),
  alpha = 0.05
)

## 4️⃣ Vytvořit MEM proměnné
if (is.null(me_obj$vectors) || ncol(as.data.frame(me_obj$vectors)) == 0) {
  
  mem_vars_df <- mem_select_df_sub |>
    select(id)
  
} else {
  
  mem_mat <- as.data.frame(me_obj$vectors)
  names(mem_mat) <- paste0("mem_", seq_len(ncol(mem_mat)))

  mem_vars_df <- mem_select_df_sub |>
    select(id) |>
    bind_cols(mem_mat)
}

## 5️⃣ Napojení zpět na panel
panel_df_mem <- panel_df_model |>
  left_join(mem_vars_df, by = "id")

## 6️⃣ Nahradit NA nulou
mem_names <- names(panel_df_mem)[grepl("^mem_", names(panel_df_mem))]

if (length(mem_names) > 0) {
  panel_df_mem <- panel_df_mem |>
    mutate(across(all_of(mem_names), ~ replace_na(., 0)))
}

mem_names
```

```{r}
panel_df_mem <- panel_df_mem |>
  mutate(
    population = replace_na(population, 0),
    pop_offset = log(population + 1)
  )
```

```{r}
train_df_mem <- panel_df_mem |> filter(year < 2025)
test_df_mem  <- panel_df_mem |> filter(year >= 2025)

train_df_zi <- panel_df_mem |> filter(year < 2025)
test_df_zi  <- panel_df_mem |> filter(year >= 2025)

train_df_base <- panel_df_model |> filter(year < 2025)
test_df_base <- panel_df_model |> filter(year >= 2025)
```

## 20. Baseline negative binomial model

```{r}
m_nb_base <- fenegbin(
  crime_count ~ ghsl_mean + clos_mean +
    n_pubs_500 + has_metro_500 + eig_mean + deg_mean + offset(pop_offset) 
    | month,
  data = train_df_base #panel_df_model
)

summary(m_nb_base)

wald(m_nb_base, "year")
wald(m_nb_base, "month")
```

```{r}
year_fe <- data.frame(
  year = names(fixef(m_nb_base)$year),
  effect = as.numeric(fixef(m_nb_base)$year)
)

month_fe <- data.frame(
  month = names(fixef(m_nb_base)$month),
  effect = as.numeric(fixef(m_nb_base)$month)
)

year_fe
month_fe
```


```{r}
ggplot(month_fe, aes(x = as.numeric(month), y = effect)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Month fixed effects",
    x = "Month",
    y = "Estimated fixed effect"
  )
```

```{r}
test_df_base$pred_nb <- predict(
  m_nb_base,
  newdata = test_df_base,
  type = "response"
)
```


```{r}
cor_base <- cor(
  log(test_df_base$crime_count + 1),
  log(test_df_base$pred_nb + 1),
  use = "complete.obs"
)
cor_base
```

```{r}
sqrt(mean(
  (test_df_base$crime_count - test_df_base$pred_nb)^2,
  na.rm = TRUE
))
```

```{r}
sqrt(mean(
  ((test_df_base$crime_count - test_df_base$pred_nb) /
   (test_df_base$pred_nb + 1))^2,
  na.rm = TRUE
))
```


```{r}
ggplot(test_df_base, aes(x = pred_nb, y = crime_count)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(
    title = "Out-of-sample prediction",
    x = "Predicted",
    y = "Observed"
  )
```



## 21. Negative binomial model with MEM filtering

```{r}
base_rhs <- c(
  "ghsl_mean",
  "betw_mean",
  "clos_mean",
  "n_pubs_500",
  "has_metro_500",
  "eig_mean",
  "deg_mean",
  "offset(pop_offset)"
)

rhs_all <- c(base_rhs, mem_names)

fml_mem <- as.formula(
  paste0(
    "crime_count ~ ",
    paste(rhs_all, collapse = " + "),
    " | year + month"
  )
)

fml_mem <- as.formula("crime_count ~ ghsl_mean + betw_mean + clos_mean + n_pubs_500 + 
    has_metro_500 + eig_mean + deg_mean + offset(pop_offset) + mem_1 | year + month")

m_nb_mem <- fenegbin(
  crime_count ~ ghsl_mean + clos_mean +
    n_pubs_500 + has_metro_500 + eig_mean + deg_mean + offset(pop_offset) + mem_1 + mem_3 + mem_5 + mem_7 + mem_8 + mem_9 + mem_10 + mem_11 + mem_12 + mem_13 + mem_14 + mem_15 + mem_17 + mem_18 + mem_19
    | month,
  data = train_df_mem
)

summary(m_nb_mem)

modelsummary(
  m_nb_mem,
  output = "img/m_nb_mem.tex"
)


```

```{r}
test_df_mem$pred_nb <- predict(
  m_nb_mem,
  newdata = test_df_mem,
  type = "response"
)
```

```{r}
cor_mem <- cor(
  log(test_df_mem$crime_count + 1),
  log(test_df_mem$pred_nb + 1),
  use = "complete.obs"
)

cor_mem
```


```{r}
sqrt(mean(
  (test_df_mem$crime_count - test_df_mem$pred_nb)^2,
  na.rm = TRUE
))
```


```{r}
sqrt(mean(
  ((test_df_mem$crime_count - test_df_mem$pred_nb) /
   (test_df_mem$pred_nb + 1))^2,
  na.rm = TRUE
))
```


## 22. Zero inflated model

```{r}
m_zinb <- zeroinfl(
  crime_count ~ ghsl_mean + clos_mean + eig_mean +
    n_pubs_500 + has_metro_500 + deg_mean + mem_1 + mem_3 + mem_5 + mem_7 + mem_8 + mem_9 + mem_10 + mem_11 + mem_12 + mem_13 + mem_14 + mem_15 + mem_17 + mem_18 + mem_19 +
    factor(year) + factor(month) + offset(pop_offset)
  |
    has_metro_500 + population,
  data = train_df_zi,
  dist = "negbin"
)

summary(m_zinb)

modelsummary(
  m_zinb,
  output = "img/zinb.tex"
)
```


```{r}
AIC(m_nb_mem)
AIC(m_zinb)
```

```{r}
BIC(m_nb_mem)
BIC(m_zinb)
```

```{r}
mean(panel_df_mem$crime_count == 0)
mean(predict(m_zinb, type = "prob")[, "0"])
```


```{r}
test_df_zi$pred_nb <- predict(
  m_nb_base,
  newdata = test_df_zi,
  type = "response"
)
```


```{r}
cor_zinb <- cor(
  log(test_df_zi$crime_count + 1),
  log(test_df_zi$pred_nb + 1),
  use = "complete.obs"
)

cor_zinb
```

```{r}
sqrt(mean(
  (test_df_zi$crime_count - test_df_zi$pred_nb)^2,
  na.rm = TRUE
))
```

```{r}
sqrt(mean(
  ((test_df_zi$crime_count - test_df_zi$pred_nb) /
   (test_df_zi$pred_nb + 1))^2,
  na.rm = TRUE
))
```




## 22. Model comparison

```{r}
etable(m_nb_base, m_nb_mem)
summary(m_zinb)
```

## 23. Moran test on residuals

```{r}
# residuals aligned to original data
train_df_base$resid_base <- residuals(
  m_nb_base,
  type = "pearson",
  na.rm = FALSE
)

# average residual by cell
resid_base_by_id <- train_df_base |>
  group_by(id) |>
  summarise(
    resid_mean = mean(resid_base, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    resid_mean = ifelse(is.nan(resid_mean), NA, resid_mean)
  ) |>
  arrange(id)

# keep only non-missing cells
keep_base <- !is.na(resid_base_by_id$resid_mean)

resid_base_clean <- resid_base_by_id[keep_base, ]

# subset neighbour structure to same cells
nb_base <- subset(nb, subset = keep_base)
lw_base <- nb2listw(nb_base, style = "W", zero.policy = TRUE)

# Moran Monte Carlo test
moran_base <- moran.mc(
  resid_base_clean$resid_mean,
  listw = lw_base,
  nsim = 999,
  zero.policy = TRUE
)

moran_base
```

```{r}
train_df_mem$resid_mem <- residuals(
  m_nb_mem,
  type = "pearson",
  na.rm = FALSE
)

resid_mem_by_id <- train_df_mem |>
  group_by(id) |>
  summarise(
    resid_mean = mean(resid_mem, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    resid_mean = ifelse(is.nan(resid_mean), NA, resid_mean)
  ) |>
  arrange(id)

keep_mem_res <- !is.na(resid_mem_by_id$resid_mean)

resid_mem_clean <- resid_mem_by_id[keep_mem_res, ]

nb_mem_res <- subset(nb, subset = keep_mem_res)
lw_mem_res <- nb2listw(nb_mem_res, style = "W", zero.policy = TRUE)

moran_mem <- moran.mc(
  resid_mem_clean$resid_mean,
  listw = lw_mem_res,
  nsim = 999,
  zero.policy = TRUE
)

moran_mem
```

## 24. Predictions

```{r}
# identify rows used in estimation
used_mem <- !is.na(residuals(m_nb_mem, type = "pearson", na.rm = FALSE))

# full-length prediction vector
train_df_mem$pred_nb_mem <- NA_real_

# fill only rows used by the model
train_df_mem$pred_nb_mem[used_mem] <- predict(m_nb_mem, type = "response")
```

```{r}
ggplot(train_df_mem, aes(x = pred_nb_mem, y = crime_count)) +
  geom_point(alpha = 0.25) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(
    title = "NB + MEM: observed vs predicted crime",
    x     = "Predicted crime (log scale)",
    y     = "Observed crime (log scale)"
  )
```

```{r}
ggplot(train_df_mem %>% filter(!is.na(pred_nb_mem)),
       aes(x = pred_nb_mem, y = crime_count)) +
  geom_hex(bins = 50) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  scale_x_log10() +
  scale_y_log10() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(
    title = "NB + MEM: observed vs predicted crime",
    x = "Predicted crime (log scale)",
    y = "Observed crime (log scale)"
  )
```

```{r}
cor(
  log(train_df_mem$crime_count + 1),
  log(train_df_mem$pred_nb_mem + 1),
  use = "complete.obs"
)
```


```{r}
sqrt(mean(
  ((train_df_mem$crime_count - train_df_mem$pred_nb_mem) /
   (train_df_mem$pred_nb_mem + 1))^2,
  na.rm = TRUE
))
```


## 25. Average observed and predicted crime by cell

```{r}
cell_avg_mem <- train_df_mem |>
  group_by(id) |>
  summarise(
    crime_avg = mean(crime_count, na.rm = TRUE),
    pred_avg  = mean(pred_nb_mem, na.rm = TRUE),
    .groups   = "drop"
  )

grid_avg_mem <- grid_full |>
  left_join(cell_avg_mem, by = "id") |>
  st_transform(st_crs(prague_wgs))
```

## 26. Maps

```{r}
edges_sf <- st_as_sf(net_prague_m, "edges")
edges_plot <- st_transform(edges_sf, st_crs(prague_wgs))
```

```{r}
ggplot(grid_avg_mem %>% filter(crime_avg > 0)) +
  geom_sf(aes(fill = crime_avg), color = NA) +
  geom_sf(
    data = edges_plot,
    color = "grey40", linewidth = 0.2, alpha = 0.7
  ) +
  scale_fill_viridis_c(
    option = "plasma",
    trans  = "log10",
    name   = "avg crime\n(log10)"
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  labs(title = "Average annual crime counts in Prague")
```

```{r}
ggplot(grid_avg_mem %>% filter(pred_avg > 0)) +
  geom_sf(aes(fill = pred_avg), color = NA) +
  geom_sf(
    data = edges_plot,
    color = "grey40", linewidth = 0.2, alpha = 0.7
  ) +
  scale_fill_viridis_c(
    option = "plasma",
    trans  = "log10",
    name   = "avg predicted\ncrime (log10)"
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  labs(title = "NB + MEM: average predicted crime per H3 cell")
```

```{r}
train_df_mem$resid_mem <- residuals(
  m_nb_mem,
  type = "pearson",
  na.rm = FALSE
)

cell_resid_pearson <- train_df_mem |>
  group_by(id) |>
  summarise(
    resid_pearson_avg = mean(resid_mem, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    resid_pearson_avg = ifelse(is.nan(resid_pearson_avg), NA, resid_pearson_avg)
  )

grid_resid_pearson <- grid_full |>
  left_join(cell_resid_pearson, by = "id") |>
  st_transform(st_crs(prague_wgs))

ggplot(grid_resid_pearson) +
  geom_sf(aes(fill = resid_pearson_avg), color = NA) +
  geom_sf(data = edges_plot,
          color = "grey40", linewidth = 0.15, alpha = 0.5) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey90",
    name = "Avg Pearson\nresidual"
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  labs(title = "Average Pearson residuals by grid cell: NB + MEM")
```

## 27. Train and test maps

```{r}

cell_avg_test <- test_df_base |>
  group_by(id) |>
  summarise(
    crime_avg = mean(crime_count, na.rm = TRUE),
    pred_avg  = mean(pred_nb, na.rm = TRUE),
    .groups   = "drop"
  )

grid_test <- grid_full |>
  left_join(cell_avg_test, by = "id") |>
  st_transform(st_crs(prague_wgs))

```


```{r}
ggplot(grid_test %>% filter(crime_avg > 0)) +
  geom_sf(aes(fill = crime_avg), color = NA) +
  geom_sf(data = edges_plot,
          color = "grey40", linewidth = 0.2, alpha = 0.7) +
  scale_fill_viridis_c(
    option = "plasma",
    trans  = "log10",
    name   = "Observed\n(log10)"
  ) +
  theme_void() +
  labs(title = "Observed crime (test set)")
```


```{r}
ggplot(grid_test %>% filter(pred_avg >= 0)) +
  geom_sf(aes(fill = pred_avg), color = NA) +
  geom_sf(data = edges_plot,
          color = "grey40", linewidth = 0.2, alpha = 0.7) +
  scale_fill_viridis_c(
    option = "plasma",
    trans  = "log10",
    name   = "Predicted\n(log10)"
  ) +
  theme_void() +
  labs(title = "Predicted crime (base NB model)")
```


```{r}

cell_avg_test_mem <- test_df_mem |>
  group_by(id) |>
  summarise(
    crime_avg = mean(crime_count, na.rm = TRUE),
    pred_avg  = mean(pred_nb, na.rm = TRUE),
    .groups   = "drop"
  )

grid_test_mem <- grid_full |>
  left_join(cell_avg_test_mem, by = "id") |>
  st_transform(st_crs(prague_wgs))

ggplot(grid_test_mem %>% filter(pred_avg > 0)) +
  geom_sf(aes(fill = pred_avg), color = NA) +
  geom_sf(data = edges_plot,
          color = "grey40", linewidth = 0.2, alpha = 0.7) +
  scale_fill_viridis_c(
    option = "plasma",
    trans  = "log10",
    name   = "Predicted\n(log10)"
  ) +
  theme_void() +
  labs(title = "Predicted crime (model with MEM filter)")
```

```{r}
cell_avg_test_zi <- test_df_zi |>
  group_by(id) |>
  summarise(
    crime_avg = mean(crime_count, na.rm = TRUE),
    pred_avg  = mean(pred_nb, na.rm = TRUE),
    .groups   = "drop"
  )

grid_test_zi <- grid_full |>
  left_join(cell_avg_test_zi, by = "id") |>
  st_transform(st_crs(prague_wgs))

ggplot(grid_test_zi %>% filter(pred_avg > 0)) +
  geom_sf(aes(fill = pred_avg), color = NA) +
  geom_sf(data = edges_plot,
          color = "grey40", linewidth = 0.2, alpha = 0.7) +
  scale_fill_viridis_c(
    option = "plasma",
    trans  = "log10",
    name   = "Predicted\n(log10)"
  ) +
  theme_void() +
  labs(title = "Predicted crime (zero inflated model)")
```

