---
title: "panel_fixed_effects"
output: html_document
date: "2025-11-21"
---

## 0. Packages

```{r}
# install.packages(c(
#   "sf", "terra", "osmdata", "dplyr", "sfnetworks", 
#   "tidygraph", "MASS", "ggplot2"
# ))

library(sf)
library(terra)
library(osmdata)
library(dplyr)
library(sfnetworks)
library(tidygraph)
library(MASS)
library(ggplot2)
library(tidyterra)
library(lubridate)
library(tidyr)
library(fixest)      # for fenegbin()
library(h3jsr)
```

## 1. Load GHSL and crimes

```{r}
# Your files
r <- rast("/Users/jachym.bielesz/diplomka/ghsl/GHS_BUILT_S_E2018_GLOBE_R2023A_54009_10_V1_0_R4_C20.tif")
names(r) <- "ghsl"   # simpler name

CRIME_FILE <- "/Users/jachym.bielesz/diplomka/kriminalita_geojson/crimes_prague.geojson"

# Crimes as sf
crimes <- st_read(CRIME_FILE)

# Make sure CRS is WGS84 (most GeoJSONs are)
if (is.na(st_crs(crimes))) {
  st_crs(crimes) <- 4326
}
```

```{r}
crimes <- crimes |>
  mutate(
    crime_time = ymd_hms(date),
    year       = year(crime_time)
  )
```


## 2. Get Prague boundary and filter crimes to Prague

```{r}
# Prague polygon
prague_poly <- getbb("Praha, Czech Republic", format_out = "polygon")

str(prague_poly, max.level = 3)

prague <- st_sfc(
  lapply(prague_poly, function(feature) {
    coords <- as.matrix(feature[[1]])
    storage.mode(coords) <- "double"
    st_polygon(list(coords))
  }),
  crs = 4326
) |>
  st_as_sf() |>
  summarise() |>
  st_make_valid()
```

```{r}
ggplot() +
  geom_sf(data = prague, fill = "lightblue", color = "black", alpha = 0.4) +
  theme_minimal() +
  labs(title = "Prague Administrative Boundary (OSM admin_level = 8)")
```

```{r}
prague_wgs <- st_transform(prague, 4326)

# Keep only crimes inside Prague
crimes_prague <- st_intersection(crimes, prague_wgs)
```


```{r}
ggplot() +
  geom_sf(data = prague_wgs,
          fill = NA,
          color = "grey60",
          linewidth = 0.4) +
  geom_sf(data = crimes_prague %>% filter(year == 2020),
          color = "red",
          size = 0.4,
          alpha = 0.4) +
  theme_void() +
  labs(title = "2020 recorded crimes in Prague")
```


## 3. Build a grid over Prague (this defines your units i)

Let’s use **500 m** cells (you can change `cellsize`).

We build it in **GHSL CRS (EPSG:54009)** so extracting raster is easy.

```{r}
# Prague in WGS84 (required by h3jsr)
prague_wgs <- st_transform(prague, 4326)

# Choose H3 resolution (tune if you want finer / coarser cells)
res_h3 <- 8

# Get H3 cells covering Prague – simple = TRUE gives a clean character vector
h3_ids <- polygon_to_cells(prague_wgs, res = res_h3, simple = TRUE)
h3_ids <- h3_ids[!is.na(h3_ids)]

# Convert H3 cells to polygons (sf, still WGS84)
hex_sf <- cell_to_polygon(h3_ids, simple = FALSE) |>
  mutate(id = seq_len(n()))

# Put hexes in GHSL CRS (54009) for raster extraction & crime join
hex_54009 <- st_transform(hex_sf, crs(r))

# Crimes in same CRS
crimes_54009 <- st_transform(crimes_prague, crs(r))

```

------------------------------------------------------------------------

## 4. Crime counts per grid cell (Y_i)

```{r}
# Which hex cell each crime falls into (still in 54009)
join_crime <- st_join(crimes_54009, hex_54009, join = st_within)

# Count crimes per hex *and year*
crime_counts_year <- join_crime |>
  st_drop_geometry() |>
  group_by(id.y, year) |>
  summarise(crime_count = n(), .groups = "drop")

# start from hex grid as our base polygon layer
grid_crime <- hex_54009 |>
  left_join(crime_counts_year, by = c("id" = "id.y")) |>
  mutate(crime_count = ifelse(is.na(crime_count), 0L, crime_count))

```

------------------------------------------------------------------------

## 5. Extract GHSL built-up per cell (B_i)

We take **mean built-up intensity** in each grid cell.

```{r}
# terra::extract with polygons
vals <- terra::extract(r, vect(grid_crime), fun = mean, na.rm = TRUE)

# `vals` has first column ID, second = ghsl mean
grid_crime$ghsl_mean <- vals$ghsl
```

------------------------------------------------------------------------

## 6. Compute street network centrality & aggregate per grid

### 6.1 Get Prague streets from OSM


```{r}

bb <- st_bbox(prague_wgs)

#car_highways <- c(
#  "motorway", "trunk", "primary", "secondary", "tertiary",
#  "motorway_link", "trunk_link", "primary_link",
#  "secondary_link", "tertiary_link",
#  "residential"
#)

car_highways <- c(
  "motorway", "trunk", "primary", "secondary", "tertiary",
  "motorway_link", "trunk_link", "primary_link",
  "secondary_link", "tertiary_link",
  "residential"
)


q_streets <- opq(bbox = bb) |>
  add_osm_feature(key = "highway", value = car_highways)

osm <- osmdata_sf(q_streets)   # this SHOULD be smaller
streets_raw <- osm$osm_lines
rm(osm); gc()
```

```{r}
# Clip to Prague and clean geometries
streets_prague <- streets_raw |> st_make_valid()
```

### 6.2 Force LINESTRING only (for sfnetwork)

```{r}
streets_clean <- streets_prague |>
  filter(st_geometry_type(geometry) %in% c("LINESTRING", "MULTILINESTRING")) |>
  st_cast("LINESTRING") |>
  filter(!st_is_empty(geometry))
```

### 6.3 Build spatial network and centrality

We compute centrality on edges (street segments). Use a metric CRS (Web Mercator, EPSG:3857):

```{r}
# Build network
net_prague <- as_sfnetwork(
  streets_clean,
  directed = FALSE,
  edges_as_lines = TRUE
)

# Project to meters
net_prague_m <- st_transform(net_prague, 3857)

# ---- NODE centralities: degree, betweenness, closeness, eigenvector ----
net_prague_m <- net_prague_m |>
  activate(nodes) |>
  mutate(
    # 1) Degree centrality (number of incident edges)
    deg  = centrality_degree(mode = "all"),
    
    # 2) Betweenness centrality (how often this node lies on shortest paths)
    betw = centrality_betweenness(),
    
    # 3) Closeness centrality (inverse of total distance to all other nodes)
    clos = centrality_closeness(),
    
    # 4) Eigenvector centrality (connected to important nodes = more important)
    eig  = centrality_eigen()
  )

# Extract nodes as sf with all four centralities
nodes_sf <- st_as_sf(net_prague_m, "nodes")
```

### 6.4 Aggregate centrality per grid cell

We need the grid in same CRS (3857), then intersect.

```{r}
# Grid in metric CRS to match the nodes (still 54009 -> 3857)
grid_3857 <- st_transform(grid_crime, 3857)

# Join nodes to grid cells they fall into
join_nodes <- st_join(grid_3857, nodes_sf, join = st_contains)

# Aggregate mean (or median, max, etc.) per cell
cent_grid_nodes <- join_nodes |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    deg_mean   = mean(deg,  na.rm = TRUE),
    betw_mean  = mean(betw, na.rm = TRUE),
    clos_mean  = mean(clos, na.rm = TRUE),
    eig_mean   = mean(eig,  na.rm = TRUE),
    .groups    = "drop"
  )

# Merge these node-based centralities back into your original grid_crime (54009)
grid_full <- grid_crime |>
  left_join(cent_grid_nodes, by = "id")
```


Now `grid_full` has, for each cell (i):

-   `crime_count` = number of crimes
-   `ghsl_mean` = mean built-up intensity
-   `betw_mean` = mean edge betweenness
-   `clos_mean` = mean node closeness

## Add pubs and metro

```{r}
## 6.5 Metro stations (within 200 m of cell) ----

bb <- st_bbox(prague_wgs)

# Metro / rail stations (you can refine this if you have better tags)
q_metro <- opq(bbox = bb) |>
  add_osm_feature(key = "railway", value = c("station", "halt", "subway_entrance"))

osm_metro <- osmdata_sf(q_metro)
metro_pts  <- osm_metro$osm_points |> st_make_valid()

# Reproject to GHSL / grid CRS (54009)
metro_54009 <- st_transform(metro_pts, crs(grid_crime))

# Join: cells with a metro within 200 m
join_metro <- st_join(
  grid_crime,
  metro_54009,
  join = st_is_within_distance,
  dist  = 500  # meters
)

# Aggregate: any metro within 200m?
metro_grid <- join_metro |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    has_metro_500 = as.integer(any(!is.na(osm_id))),  # 1 if >=1 station
    .groups = "drop"
  )

```


```{r}
## 6.6 Pubs / bars (count within 200 m of cell) ----

q_pubs <- opq(bbox = bb) |>
  add_osm_feature(
    key   = "amenity",
    value = c("pub", "bar", "restaurant", "cafe")
  )

osm_pubs <- osmdata_sf(q_pubs)
pub_pts  <- osm_pubs$osm_points |> st_make_valid()

pubs_54009 <- st_transform(pub_pts, crs(grid_crime))

join_pubs <- st_join(
  grid_crime,
  pubs_54009,
  join = st_is_within_distance,
  dist  = 500  # meters
)

pubs_grid <- join_pubs |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    n_pubs_500 = sum(!is.na(osm_id)),          # number of pubs/bars/etc within 200m
    .groups    = "drop"
  )

```

```{r}
## 6.7 Attach pubs & metro to grid_full ----

grid_full <- grid_full |>
  left_join(pubs_grid,  by = "id") |>
  left_join(metro_grid, by = "id") |>
  mutate(
    n_pubs_500    = ifelse(is.na(n_pubs_500), 0L, n_pubs_500),
    has_metro_500 = ifelse(is.na(has_metro_500), 0L, has_metro_500)
  )

```


```{r}
static_vars <- grid_full |>
  st_drop_geometry() |>
  select(id,
         ghsl_mean,
         betw_mean,
         clos_mean,
         n_pubs_500,
         has_metro_500,
         eig_mean,
         deg_mean)

```




## 7. Prepare data frame for regression

```{r}
ids   <- sort(unique(static_vars$id))
years <- sort(unique(crime_counts_year$year))

panel_base <- expand_grid(id = ids, year = years)

panel_df <- panel_base |>
  left_join(crime_counts_year,
            by = c("id" = "id.y", "year" = "year")) |>
  mutate(
    crime_count = ifelse(is.na(crime_count), 0L, crime_count)
  ) |>
  left_join(static_vars, by = "id")

```


```{r}
panel_df_model <- panel_df |>
  filter(
    !is.na(ghsl_mean),
    !is.na(betw_mean),
    !is.na(clos_mean),
    !is.na(n_pubs_500),
    !is.na(eig_mean),
    !is.na(deg_mean),
    !is.na(has_metro_500)
  )

```



------------------------------------------------------------------------

## 8. Check distribution & choose model

Quick diagnostics:

```{r}
mean(panel_df_model$crime_count)
var(panel_df_model$crime_count)
mean(panel_df_model$crime_count == 0)  # share of zeros
```

`var >> mean` → use **Negative Binomial**.


## 9. Run models


```{r}
# two-way fixed effects NB: id FE + year FE
m_nb_fe <- fenegbin(
  crime_count ~ ghsl_mean + betw_mean + clos_mean +
    n_pubs_500 + has_metro_500 + eig_mean + deg_mean | year,
  data = panel_df_model
)

summary(m_nb_fe)


```


## Graphs

```{r}
# Predictions on original scale (expected crime count)
panel_df_model$pred_nb <- predict(m_nb_fe, type = "response")

```


```{r}
library(ggplot2)

ggplot(panel_df_model, aes(x = pred_nb, y = crime_count)) +
  geom_point(alpha = 0.25) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(
    title = "Panel NB FE: observed vs predicted crime",
    x     = "Predicted crime (log scale)",
    y     = "Observed crime (log scale)"
  )

```


```{r}
library(dplyr)

cell_avg <- panel_df_model |>
  group_by(id) |>
  summarise(
    crime_avg = mean(crime_count),
    pred_avg  = mean(pred_nb),
    .groups   = "drop"
  )

grid_avg <- grid_full |>
  left_join(cell_avg, by = "id") |>
  st_transform(st_crs(prague_wgs))

```

```{r}
edges_sf <- st_as_sf(net_prague_m, "edges")
edges_plot <- st_transform(edges_sf, st_crs(prague_wgs))
grid_plot <- st_transform(grid_avg, st_crs(prague_wgs))

library(dplyr)
library(sf)
library(ggplot2)

# 1️⃣ Average crime count per cell across all years
crime_avg <- crime_counts_year |>
  group_by(id.y) |>
  summarise(
    crime_avg = mean(crime_count, na.rm = TRUE),
    .groups = "drop"
  )

# 2️⃣ Join with your grid
grid_crime_avg <- hex_54009 |>
  left_join(crime_avg, by = c("id" = "id.y")) |>
  mutate(
    crime_avg = ifelse(is.na(crime_avg), 0, crime_avg)
  )

# 3️⃣ Reproject to match Prague outline CRS
grid_plot <- st_transform(grid_crime_avg, st_crs(prague_wgs))

# 4️⃣ Plot (only cells with nonzero averages)
grid_positive <- grid_plot |> filter(crime_avg > 0)

ggplot() +
  geom_sf(data = grid_positive,
          aes(fill = crime_avg),
          color = NA) +
  geom_sf(data = edges_plot,
          color = "grey40", linewidth = 0.2, alpha = 0.7) +
  scale_fill_viridis_c(
    option = "plasma",
    trans  = "log10",
    name   = "avg crime count\n(log10)"
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(title = "Average annual crime counts in Prague (500 m grid)")
```



```{r}

ggplot(grid_avg %>% filter(pred_avg > 10) ) +
  geom_sf(aes(fill = pred_avg), color = NA) +
  geom_sf(data = edges_plot,
          color = "grey40", linewidth = 0.2, alpha = 0.7) +
  scale_fill_viridis_c(
    option = "plasma",
    trans  = "log10",
    name   = "avg predicted\ncrime (log)",
    na.value = NA
  ) +
  theme_void() +
  labs(
    title = "Panel NB FE: average predicted crime per 500 m cell"
  )


```


