---
title: "spatial_durbin_one_year_h3"
output: html_document
date: "2025-11-22"
---

## 0. Packages

```{r}
# install.packages(c(
#   "sf", "terra", "osmdata", "dplyr", "sfnetworks", 
#   "tidygraph", "MASS", "ggplot2"
# ))

library(sf)
library(terra)
library(osmdata)
library(dplyr)
library(sfnetworks)
library(tidygraph)
library(MASS)
library(ggplot2)
library(tidyterra)
library(lubridate)
library(h3jsr)
```

## 1. Load GHSL and crimes

```{r}
# Your files
r <- rast("/Users/jachym.bielesz/diplomka/ghsl/GHS_BUILT_S_E2018_GLOBE_R2023A_54009_10_V1_0_R4_C20.tif")
names(r) <- "ghsl"   # simpler name

CRIME_FILE <- "/Users/jachym.bielesz/diplomka/kriminalita_geojson/crimes_prague.geojson"

# Crimes as sf
crimes <- st_read(CRIME_FILE)

# Make sure CRS is WGS84 (most GeoJSONs are)
if (is.na(st_crs(crimes))) {
  st_crs(crimes) <- 4326
}
```

```{r}
# Add year and keep only 2025 crimes
crimes_2025 <- crimes |>
  mutate(
    crime_time = ymd_hms(date),   # <<< CHANGE THIS COLUMN NAME
    year       = year(crime_time)
  ) |>
  filter(year == 2025)
```


## 2. Get Prague boundary and filter crimes to Prague

```{r}
# Prague admin boundary from OSM
prg_query <- opq("Prague, Czechia") |>
  add_osm_feature(key = "boundary", value = "administrative") |>
  add_osm_feature(key = "admin_level", value = "8")

prg_osm <- osmdata_sf(prg_query)

prague <- prg_osm$osm_multipolygons |>
  dplyr::filter(name %in% c("Praha", "Prague")) |>
  slice(1) |>
  st_make_valid()
```

```{r}
ggplot() +
  geom_sf(data = prague, fill = "lightblue", color = "black", alpha = 0.4) +
  theme_minimal() +
  labs(title = "Prague Administrative Boundary (OSM admin_level = 8)")
```

```{r}
prague_wgs <- st_transform(prague, 4326)

# Keep only crimes inside Prague
crimes_prague <- st_intersection(crimes_2025, prague_wgs)
```

## 3. Build a grid over Prague (this defines your units i)

Let’s use **500 m** cells (you can change `cellsize`).

We build it in **GHSL CRS (EPSG:54009)** so extracting raster is easy.

```{r}
# Prague must be WGS84 for h3jsr
prague_wgs <- st_transform(prague, 4326)

# Choose H3 resolution (try 8 first; you can tweak)
res_h3 <- 8

# Get all H3 cells covering Prague – simple = TRUE gives a clean character vector
h3_ids <- polygon_to_cells(prague_wgs, res = res_h3, simple = TRUE)
h3_ids <- h3_ids[!is.na(h3_ids)]   # just in case

# Convert H3 cells to polygons (sf, in WGS84)
hex_sf <- cell_to_polygon(h3_ids, simple = FALSE)
hex_sf <- hex_sf |>
  mutate(id = seq_len(n()))

# Put hexes in GHSL CRS (54009) for raster extraction & crime join
hex_54009 <- st_transform(hex_sf, crs(r))

# Crimes in same CRS
crimes_54009 <- st_transform(crimes_prague, crs(r))

```

------------------------------------------------------------------------

## 4. Crime counts per grid cell (Y_i)

```{r}
# Which hex cell each crime falls into
join_crime <- st_join(crimes_54009, hex_54009, join = st_within)

# Count crimes per hex cell
crime_counts <- join_crime |>
  st_drop_geometry() |>
  group_by(id.y) |>
  summarise(crime_count = n(), .groups = "drop")

# Merge counts back to hex polygons
grid_crime <- hex_54009 |>
  left_join(crime_counts,by = c("id" = "id.y")) |>
  mutate(crime_count = ifelse(is.na(crime_count), 0L, crime_count))

```

------------------------------------------------------------------------

## 5. Extract GHSL built-up per cell (B_i)

We take **mean built-up intensity** in each grid cell.

```{r}
# terra::extract with polygons
vals <- terra::extract(r, vect(grid_crime), fun = mean, na.rm = TRUE)

# `vals` has first column ID, second = ghsl mean
grid_crime$ghsl_mean <- vals$ghsl
```

------------------------------------------------------------------------

## 6. Compute street network centrality & aggregate per grid

### 6.1 Get Prague streets from OSM


```{r}

bb <- st_bbox(prague_wgs)

#car_highways <- c(
#  "motorway", "trunk", "primary", "secondary", "tertiary",
#  "motorway_link", "trunk_link", "primary_link",
#  "secondary_link", "tertiary_link",
#  "residential"
#)

car_highways <- c(
  "footway", "pedestrian", "path", "living_street",
  "track", "residential", "steps"
)


q_streets <- opq(bbox = bb) |>
  add_osm_feature(key = "highway", value = car_highways)

osm <- osmdata_sf(q_streets)   # this SHOULD be smaller
streets_raw <- osm$osm_lines
rm(osm); gc()
```

```{r}
# Clip to Prague and clean geometries
streets_prague <- streets_raw |> st_make_valid()
```

### 6.2 Force LINESTRING only (for sfnetwork)

```{r}
streets_clean <- streets_prague |>
  filter(st_geometry_type(geometry) %in% c("LINESTRING", "MULTILINESTRING")) |>
  st_cast("LINESTRING") |>
  filter(!st_is_empty(geometry))
```

### 6.3 Build spatial network and centrality

We compute centrality on edges (street segments). Use a metric CRS (Web Mercator, EPSG:3857):

```{r}
# Build network
net_prague <- as_sfnetwork(
  streets_clean,
  directed = FALSE,
  edges_as_lines = TRUE
)

# Project to meters
net_prague_m <- st_transform(net_prague, 3857)

# ---- NODE centralities: degree, betweenness, closeness, eigenvector ----
net_prague_m <- net_prague_m |>
  activate(nodes) |>
  mutate(
    # 1) Degree centrality (number of incident edges)
    deg  = centrality_degree(mode = "all"),
    
    # 2) Betweenness centrality (how often this node lies on shortest paths)
    betw = centrality_betweenness(),
    
    # 3) Closeness centrality (inverse of total distance to all other nodes)
    clos = centrality_closeness(),
    
    # 4) Eigenvector centrality (connected to important nodes = more important)
    eig  = centrality_eigen()
  )

# Extract nodes as sf with all four centralities
nodes_sf <- st_as_sf(net_prague_m, "nodes")

```

### 6.4 Aggregate centrality per grid cell

We need the grid in same CRS (3857), then intersect.

```{r}
# Grid in metric CRS to match the nodes (still 54009 -> 3857)
grid_3857 <- st_transform(grid_crime, 3857)

# Join nodes to grid cells they fall into
join_nodes <- st_join(grid_3857, nodes_sf, join = st_contains)

# Aggregate mean (or median, max, etc.) per cell
cent_grid_nodes <- join_nodes |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    deg_mean   = mean(deg,  na.rm = TRUE),
    betw_mean  = mean(betw, na.rm = TRUE),
    clos_mean  = mean(clos, na.rm = TRUE),
    eig_mean   = mean(eig,  na.rm = TRUE),
    .groups    = "drop"
  )

# Merge these node-based centralities back into your original grid_crime (54009)
grid_full <- grid_crime |>
  left_join(cent_grid_nodes, by = "id")

```

Now `grid_full` has, for each cell (i):

-   `crime_count` = number of crimes
-   `ghsl_mean` = mean built-up intensity
-   `betw_mean` = mean edge betweenness
-   `clos_mean` = mean node closeness

## Add pubs and metro

```{r}
## 6.5 Metro stations (within 200 m of cell) ----

bb <- st_bbox(prague_wgs)

# Metro / rail stations (you can refine this if you have better tags)
q_metro <- opq(bbox = bb) |>
  add_osm_feature(key = "railway", value = c("station", "halt", "subway_entrance"))

osm_metro <- osmdata_sf(q_metro)
metro_pts  <- osm_metro$osm_points |> st_make_valid()

# Reproject to GHSL / grid CRS (54009)
metro_54009 <- st_transform(metro_pts, crs(grid_crime))

# Join: cells with a metro within 200 m
join_metro <- st_join(
  grid_crime,
  metro_54009,
  join = st_is_within_distance,
  dist  = 500  # meters
)

# Aggregate: any metro within 200m?
metro_grid <- join_metro |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    has_metro_500 = as.integer(any(!is.na(osm_id))),  # 1 if >=1 station
    .groups = "drop"
  )

```


```{r}
## 6.6 Pubs / bars (count within 200 m of cell) ----

q_pubs <- opq(bbox = bb) |>
  add_osm_feature(
    key   = "amenity",
    value = c("pub", "bar", "restaurant", "cafe")
  )

osm_pubs <- osmdata_sf(q_pubs)
pub_pts  <- osm_pubs$osm_points |> st_make_valid()

pubs_54009 <- st_transform(pub_pts, crs(grid_crime))

join_pubs <- st_join(
  grid_crime,
  pubs_54009,
  join = st_is_within_distance,
  dist  = 500  # meters
)

pubs_grid <- join_pubs |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    n_pubs_500 = sum(!is.na(osm_id)),          # number of pubs/bars/etc within 200m
    .groups    = "drop"
  )

```

```{r}
## 6.7 Attach pubs & metro to grid_full ----

grid_full <- grid_full |>
  left_join(pubs_grid,  by = "id") |>
  left_join(metro_grid, by = "id") |>
  mutate(
    n_pubs_500    = ifelse(is.na(n_pubs_500), 0L, n_pubs_500),
    has_metro_500 = ifelse(is.na(has_metro_500), 0L, has_metro_500)
  )

```




## 7. Prepare data frame for regression

```{r}
grid_sdm_2025 <- grid_full |>
  mutate(
    log_crime = log1p(crime_count),
    ghsl_s    = as.numeric(scale(ghsl_mean)),
    betw_s    = as.numeric(scale(betw_mean)),
    clos_s    = as.numeric(scale(clos_mean)),
    eig_s    = as.numeric(scale(eig_mean)),
    deg_s    = as.numeric(scale(deg_mean)),
    pubs_s    = as.numeric(scale(n_pubs_500))
  ) |>
  filter(!is.na(log_crime), !is.na(ghsl_s), !is.na(betw_s), !is.na(eig_mean),  
         !is.na(deg_mean),
         !is.na(clos_s), !is.na(pubs_s), !is.na(has_metro_500))

df <- grid_full |>
  st_drop_geometry() |>
  mutate(
    crime_any = as.integer(crime_count > 0)
  )
```

------------------------------------------------------------------------

## 8. Check distribution & choose model

Quick diagnostics:

```{r}
mean(df$crime_count)
var(df$crime_count)
mean(df$crime_count == 0)  # share of zeros
```

`var >> mean` → use **Negative Binomial**.


## 9. Run models


```{r}
## 9. Spatial Durbin model for 2025 -----------------------------

library(spdep)
library(spatialreg)

# 9.1 Spatial weights (Queen contiguity on the grid)
nb_2025 <- poly2nb(grid_sdm_2025)
lw_2025 <- nb2listw(nb_2025, style = "W", zero.policy = TRUE)

# 9.2 Spatial Durbin Model (Gaussian – classical SDM)
sdm_2025 <- lagsarlm(
  log_crime ~ ghsl_s + clos_s + betw_s + has_metro_500 + n_pubs_500 + eig_s + deg_s,
  data  = grid_sdm_2025,
  listw = lw_2025,
  type  = "mixed",
  zero.policy = TRUE
)

summary(sdm_2025)

```


```{r}
imp_2025 <- impacts(
  sdm_2025,
  listw = lw_2025,
  R = 500,        # bootstrap
  zstats = TRUE
)

summary(imp_2025)

```

## Pretty images

```{r}

library(ggplot2)
library(sf)
library(dplyr)

# edges in same CRS as Prague outline (WGS84)
edges_plot <- st_transform(edges_sf, st_crs(prague_wgs))

# optional: log-transform for prettier colour scale
edges_plot <- edges_plot |>
  mutate(betw_log = log1p(betw))    # log(1 + betweenness)

ggplot() +
  geom_sf(data = prague_wgs,
          fill = NA, color = "grey70", linewidth = 0.4) +
  geom_sf(data = edges_plot,
          aes(color = betw_log),
          linewidth = 0.5, alpha = 0.9) +
  scale_color_viridis_c(
    option = "magma",
    name   = "betweenness\n(log1p)",
    guide  = guide_colorbar(barheight = unit(3, "cm"))
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(title = "Street betweenness centrality in Prague")


```


```{r}
# threshold for "high betweenness" (top 1%)
thr <- quantile(edges_plot$betw, 0.99, na.rm = TRUE)

edges_top <- edges_plot |>
  filter(betw >= thr)

ggplot() +
  geom_sf(data = edges_plot,
          color = "grey80", linewidth = 0.2) +
  geom_sf(data = edges_top,
          color = "red3", linewidth = 1.1) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Most central streets in Prague (top 1% betweenness)"
  )

```


```{r}
grid_plot <- st_transform(grid_full, st_crs(prague_wgs))

grid_positive <- grid_plot |>
  filter(crime_count > 0)

nodes_plot <- st_transform(nodes_sf, st_crs(prague_wgs))

nodes_plot <- nodes_plot %>% filter(betw>0) |>
  mutate(
    betw_log = log1p(betw),
    clos_log = log1p(clos),
    deg_log  = log1p(deg),
    eig_log  = log1p(eig)
  )

ggplot() +
  geom_sf(data = grid_positive,
          aes(fill = crime_count),
          color = NA) +
  geom_sf(data = nodes_plot,
          color = "grey40", linewidth = 0.2, alpha = 0.7) +
  scale_fill_viridis_c(
    option = "plasma",
    trans  = "log10",                    # <<< LOG SCALE
    breaks = c(1, 10, 100, 1000, 10000),
    name   = "crime count\n(log10)"
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(title = "Crime counts (> 0) in Prague, log scale (500 m grid)")


```

```{r}
# predicted log-crime from SDM
grid_sdm_2025$pred_log <- as.numeric(predict(sdm_2025))

# back-transform to crime scale: log1p => exp - 1
grid_sdm_2025$pred_crime <- exp(grid_sdm_2025$pred_log) - 1

```


```{r}
ggplot(grid_sdm_2025, aes(x = pred_crime, y = crime_count)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(
    title = "SDM 2025: observed vs predicted crime",
    x     = "Predicted crime (log scale)",
    y     = "Observed crime (log scale)"
  )

```

```{r}
# Project to meters
net_prague_m <- st_transform(net_prague, 3857)

# Edge centrality
net_prague_m <- net_prague_m |>
  activate(edges) |>
  mutate(
    betw = centrality_edge_betweenness(),
    len  = st_length(geometry)
  )

# Node closeness
net_prague_m <- net_prague_m |>
  activate(nodes) |>
  mutate(clos = centrality_closeness())

# Extract node closeness as a vector
node_clos <- net_prague_m %>%
  activate(nodes) %>%
  pull(clos)

# Push closeness to edges
net_prague_m <- net_prague_m |>
  activate(edges) |>
  mutate(
    clos_edge = (node_clos[from] + node_clos[to]) / 2
  )

edges_sf <- st_as_sf(net_prague_m, "edges")
edges_plot <- st_transform(edges_sf, st_crs(prague_wgs))

grid_sdm_2025_plot <- st_transform(grid_sdm_2025, st_crs(prague_wgs))

ggplot(grid_sdm_2025_plot |>
  filter(crime_count > 0)) +
  geom_sf(aes(fill = pred_crime), color = NA) +
  geom_sf(data = edges_plot,
          color = "grey40", linewidth = 0.2, alpha = 0.7) +
  scale_fill_viridis_c(
    trans  = "log10",
    name   = "predicted\ncrime (2025)",
    option = "plasma",
    na.value = NA,
    breaks = c(1, 10, 100, 1000, 10000),
  ) +
  theme_void() +
  labs(
    title = "Spatial Durbin model (SDM): predicted crime in 2025"
  )

```


```{r}
library(ggplot2)
library(sf)
library(dplyr)

# nodes in same CRS as Prague outline (WGS84)
nodes_plot <- st_transform(nodes_sf, st_crs(prague_wgs))

# log-transform for smoother color gradient (optional)
nodes_plot <- nodes_plot %>% filter(betw>0) |>
  mutate(
    betw_log = log1p(betw),
    clos_log = log1p(clos),
    deg_log  = log1p(deg),
    eig_log  = log1p(eig)
  )

# ---- Example 1: Betweenness centrality map ----
ggplot() +
  geom_sf(data = grid_positive,
          fill = NA, color = "grey70", linewidth = 0.4) +
  geom_sf(data = nodes_plot,
          aes(color = betw_log),
          size = 0.8, alpha = 0.8) +
  scale_color_viridis_c(
    option = "magma",
    name   = "Node betweenness\n(log1p)",
    guide  = guide_colorbar(barheight = unit(3, "cm"))
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(title = "Node betweenness centrality in Prague")

# ---- Example 2: Highlight top 1% most central intersections ----
thr <- quantile(nodes_plot$betw, 0.99, na.rm = TRUE)

nodes_top <- nodes_plot |> filter(betw >= thr)

ggplot() +
  geom_sf(data = nodes_plot,
          color = "grey85", size = 0.4, alpha = 0.4) +
  geom_sf(data = nodes_top,
          color = "red3", size = 1.2, alpha = 0.9) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Most central intersections in Prague (top 1% betweenness)")
```

