---
title: "testing_regression"
output: html_notebook
---

## 0. Packages

```{r}
# install.packages(c(
#   "sf", "terra", "osmdata", "dplyr", "sfnetworks", 
#   "tidygraph", "MASS", "ggplot2"
# ))

library(sf)
library(terra)
library(osmdata)
library(dplyr)
library(sfnetworks)
library(tidygraph)
library(MASS)
library(ggplot2)
library(tidyterra)
```

## 1. Load GHSL and crimes

```{r}
# Your files
r <- rast("/Users/jachym.bielesz/Downloads/GHS_BUILT_S_E2018_GLOBE_R2023A_54009_10_V1_0_R4_C20/GHS_BUILT_S_E2018_GLOBE_R2023A_54009_10_V1_0_R4_C20.tif")
names(r) <- "ghsl"   # simpler name

CRIME_FILE <- "/Users/jachym.bielesz/diplomka/kriminalita_geojson/crimes_prague.geojson"

# Crimes as sf
crimes <- st_read(CRIME_FILE)

# Make sure CRS is WGS84 (most GeoJSONs are)
if (is.na(st_crs(crimes))) {
  st_crs(crimes) <- 4326
}
```

## 2. Get Prague boundary and filter crimes to Prague

```{r}
# Prague admin boundary from OSM
prg_query <- opq("Prague, Czechia") |>
  add_osm_feature(key = "boundary", value = "administrative") |>
  add_osm_feature(key = "admin_level", value = "8")

prg_osm <- osmdata_sf(prg_query)

prague <- prg_osm$osm_multipolygons |>
  dplyr::filter(name %in% c("Praha", "Prague")) |>
  slice(1) |>
  st_make_valid()
```

```{r}
ggplot() +
  geom_sf(data = prague, fill = "lightblue", color = "black", alpha = 0.4) +
  theme_minimal() +
  labs(title = "Prague Administrative Boundary (OSM admin_level = 8)")
```

```{r}
prague_wgs <- st_transform(prague, 4326)

# Keep only crimes inside Prague
crimes_prague <- st_intersection(crimes, prague_wgs)
```

## 3. Build a grid over Prague (this defines your units i)

Let’s use **500 m** cells (you can change `cellsize`).

We build it in **GHSL CRS (EPSG:54009)** so extracting raster is easy.

```{r}
# Prague in GHSL CRS
prague_54009 <- st_transform(prague_wgs, crs(r))

# Grid (500 m cells)
grid_54009 <- st_make_grid(prague_54009, cellsize = 500, square = TRUE)
grid_54009 <- st_sf(id = seq_along(grid_54009), geometry = grid_54009)

# Crimes in same CRS
crimes_54009 <- st_transform(crimes_prague, crs(r))
```

------------------------------------------------------------------------

## 4. Crime counts per grid cell (Y_i)

```{r}
# Which grid cell each crime falls into
join_crime <- st_join(crimes_54009, grid_54009, join = st_within)

# Count crimes per grid cell
crime_counts <- join_crime |>
  st_drop_geometry() |>
  group_by(id.y) |>
  summarise(crime_count = n(), .groups = "drop")

# Merge counts back to grid
grid_crime <- grid_54009 |>
  left_join(crime_counts, by = c("id" = "id.y")) |>
  mutate(crime_count = ifelse(is.na(crime_count), 0L, crime_count))
```

------------------------------------------------------------------------

## 5. Extract GHSL built-up per cell (B_i)

We take **mean built-up intensity** in each grid cell.

```{r}
# terra::extract with polygons
vals <- terra::extract(r, vect(grid_crime), fun = mean, na.rm = TRUE)

# `vals` has first column ID, second = ghsl mean
grid_crime$ghsl_mean <- vals$ghsl
```

------------------------------------------------------------------------

## 6. Compute street network centrality & aggregate per grid

### 6.1 Get Prague streets from OSM

```{r}

bb <- st_bbox(prague_wgs)

car_highways <- c(
  "motorway", "trunk", "primary", "secondary", "tertiary",
  "motorway_link", "trunk_link", "primary_link",
  "secondary_link", "tertiary_link",
  "residential"
  # I dropped "service" and "unclassified" here to shrink data
)

q_streets <- opq(bbox = bb) |>
  add_osm_feature(key = "highway", value = car_highways)

osm <- osmdata_sf(q_streets)   # this SHOULD be smaller
streets_raw <- osm$osm_lines
rm(osm); gc()
```

```{r}
# Clip to Prague and clean geometries
streets_prague <- streets_raw |> st_make_valid()
```

### 6.2 Force LINESTRING only (for sfnetwork)

```{r}
streets_clean <- streets_prague |>
  filter(st_geometry_type(geometry) %in% c("LINESTRING", "MULTILINESTRING")) |>
  st_cast("LINESTRING") |>
  filter(!st_is_empty(geometry))
```

### 6.3 Build spatial network and centrality

We compute centrality on edges (street segments). Use a metric CRS (Web Mercator, EPSG:3857):

```{r}
# Build network
net_prague <- as_sfnetwork(
  streets_clean,
  directed = FALSE,
  edges_as_lines = TRUE
)

# Project to meters
net_prague_m <- st_transform(net_prague, 3857)

# Edge centrality
net_prague_m <- net_prague_m |>
  activate(edges) |>
  mutate(
    betw = centrality_edge_betweenness(),
    len  = st_length(geometry)
  )

# Node closeness
net_prague_m <- net_prague_m |>
  activate(nodes) |>
  mutate(clos = centrality_closeness())

# Extract node closeness as a vector
node_clos <- net_prague_m %>%
  activate(nodes) %>%
  pull(clos)

# Push closeness to edges
net_prague_m <- net_prague_m |>
  activate(edges) |>
  mutate(
    clos_edge = (node_clos[from] + node_clos[to]) / 2
  )
```

### 6.4 Aggregate centrality per grid cell

We need the grid in same CRS (3857), then intersect.

```{r}
## 5) Edges as sf with BOTH centralities ----
edges_sf <- st_as_sf(net_prague_m, "edges")
# edges_sf is already in 3857 (same as net_prague_m), no need to reproject again

## 6) Reproject grid to 3857 to match edges ----
grid_3857 <- st_transform(grid_crime, 3857)

## 7) Join edges to grid cells they intersect ----
join_cent <- st_join(grid_3857, edges_sf, join = st_intersects)

## 8) Aggregate mean centrality per cell ----
cent_grid <- join_cent |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    betw_mean = mean(betw, na.rm = TRUE),
    clos_mean = mean(clos_edge, na.rm = TRUE),
    .groups   = "drop"
  )

## 9) Merge centrality back into original grid_crime (54009) 
grid_full <- grid_crime |>
  left_join(cent_grid, by = "id")
```

Now `grid_full` has, for each cell (i):

-   `crime_count` = number of crimes
-   `ghsl_mean` = mean built-up intensity
-   `betw_mean` = mean edge betweenness
-   `clos_mean` = mean node closeness

## Add pubs and metro

```{r}
## 6.5 Metro stations (within 200 m of cell) ----

bb <- st_bbox(prague_wgs)

# Metro / rail stations (you can refine this if you have better tags)
q_metro <- opq(bbox = bb) |>
  add_osm_feature(key = "railway", value = c("station", "halt", "subway_entrance"))

osm_metro <- osmdata_sf(q_metro)
metro_pts  <- osm_metro$osm_points |> st_make_valid()

# Reproject to GHSL / grid CRS (54009)
metro_54009 <- st_transform(metro_pts, crs(grid_crime))

# Join: cells with a metro within 200 m
join_metro <- st_join(
  grid_crime,
  metro_54009,
  join = st_is_within_distance,
  dist  = 200  # meters
)

# Aggregate: any metro within 200m?
metro_grid <- join_metro |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    has_metro_200 = as.integer(any(!is.na(osm_id))),  # 1 if >=1 station
    .groups = "drop"
  )

```


```{r}
## 6.6 Pubs / bars (count within 200 m of cell) ----

q_pubs <- opq(bbox = bb) |>
  add_osm_feature(
    key   = "amenity",
    value = c("pub", "bar", "restaurant", "cafe")
  )

osm_pubs <- osmdata_sf(q_pubs)
pub_pts  <- osm_pubs$osm_points |> st_make_valid()

pubs_54009 <- st_transform(pub_pts, crs(grid_crime))

join_pubs <- st_join(
  grid_crime,
  pubs_54009,
  join = st_is_within_distance,
  dist  = 200  # meters
)

pubs_grid <- join_pubs |>
  st_drop_geometry() |>
  group_by(id) |>
  summarise(
    n_pubs_200 = sum(!is.na(osm_id)),          # number of pubs/bars/etc within 200m
    .groups    = "drop"
  )

```

```{r}
## 6.7 Attach pubs & metro to grid_full ----

grid_full <- grid_full |>
  left_join(pubs_grid,  by = "id") |>
  left_join(metro_grid, by = "id") |>
  mutate(
    n_pubs_200    = ifelse(is.na(n_pubs_200), 0L, n_pubs_200),
    has_metro_200 = ifelse(is.na(has_metro_200), 0L, has_metro_200)
  )

```




## 7. Prepare data frame for regression

```{r}
df <- grid_full |>
  st_drop_geometry() |>
  mutate(
    crime_any = as.integer(crime_count > 0)
  )
```

------------------------------------------------------------------------

## 8. Check distribution & choose model

Quick diagnostics:

```{r}
mean(df$crime_count)
var(df$crime_count)
mean(df$crime_count == 0)  # share of zeros
```

You’ll probably see: `var >> mean` → use **Negative Binomial**.

------------------------------------------------------------------------

## 9. Run models

### 9.1 Negative Binomial (crime frequency)

```{r}
# Remove rows with all NA covariates (e.g. where no streets intersect)
df_nb <- df |>
  filter(!is.na(betw_mean) & !is.na(clos_mean) & !is.na(ghsl_mean))
  #filter(!is.na(betw_mean) & !is.na(ghsl_mean))

model_nb <- glm.nb(
  crime_count ~ ghsl_mean + betw_mean + clos_mean,
  data = df_nb
)

summary(model_nb)
```

Interpretation:

-   `exp(coef(model_nb)["ghsl_mean"])` = multiplicative effect on crime count for 1-unit increase in GHSL built-up.
-   Same for `betw_mean` and `clos_mean`.

### 9.2 Logistic (crime happens vs not)

```{r}
df_logit <- df_nb  # same filtered set

model_logit <- glm(
  crime_any ~ ghsl_mean + betw_mean + clos_mean,
  data = df_logit,
  family = binomial
)

summary(model_logit)
```

Interpretation:

-   `exp(coef(model_logit)["ghsl_mean"])` = odds ratio of *having at least one crime* in the cell per unit GHSL increase.
-   Likewise for centrality.

## 10. Quick maps to see patterns

```{r}
# Crime rate
grid_plot <- grid_full

ggplot() +
  geom_sf(data = grid_plot, aes(fill = crime_count), color = NA) +
  scale_fill_viridis_c(na.value = NA) +
  theme_void() +
  labs(title = "Crime counts per 500m cell")

ggplot() +
  geom_sf(data = grid_plot, aes(fill = ghsl_mean), color = NA) +
  scale_fill_viridis_c(na.value = NA) +
  theme_void() +
  labs(title = "GHSL built-up per cell")
```

## 11 Visualization

### Effect plot for GHSL mean

```{r}

ggplot(df_nb, aes(ghsl_mean, crime_count)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "glm", method.args = list(family = "nb"), color = "red") +
  theme_minimal() +
  labs(title = "Effect of Built-Up Density (GHSL) on Crime",
       x = "GHSL Built-up Mean",
       y = "Crime Count")


```

### Effect plot for Betweenness

```{r}
ggplot(df_nb, aes(betw_mean, crime_count)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "glm", method.args = list(family = "nb"), color = "blue") +
  scale_x_log10() +   # betweenness is skewed
  theme_minimal() +
  labs(title = "Effect of Street Betweenness on Crime",
       x = "Mean Betweenness (log-scale)",
       y = "Crime Count")

```

### Partial dependence / marginal effects

```{r}
library(effects)

plot(allEffects(model_nb))

```

### Residual spatial map

```{r}
df_nb$resid <- residuals(model_nb, type = "pearson")


# Join residuals back to the spatial grid
grid_resid <- grid_full |>
  left_join(df_nb |> dplyr::select(id, resid), by = "id")

# Now plot using the sf object with geometry
ggplot(grid_resid) +
  geom_sf(aes(fill = resid), color = NA) +
  scale_fill_gradient2() +
  theme_minimal() +
  labs(title = "Spatial Residuals (Negative Binomial Model)",
       fill  = "Pearson residual")


```

### Predicted vs observed crimes

```{r}
df_nb$pred <- predict(model_nb, type = "response")

ggplot(df_nb, aes(pred, crime_count)) +
  geom_point(alpha = 0.5) +
  geom_abline(color = "red") +
  theme_minimal() +
  labs(title = "Predicted vs. Observed Crime",
       x = "Predicted",
       y = "Observed")

```

### Predicted crime

```{r}
df_nb$pred <- predict(model_nb, type = "response")

grid_pred <- grid_full |>
  left_join(df_nb |> dplyr::select(id, pred), by = "id")

ggplot(grid_pred) +
  geom_sf(aes(fill = pred), color = NA) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Predicted Crime Counts (NB Model)",
       fill  = "Predicted\ncrime count")

```
